\pkg{dataMaid}: your assistant for documenting supervised data quality screening in \proglang{R}

### 2017-11-08 latex/article_vol3.tex
First round of tests

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should generally not
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the information needed for identifying issues.

But even when tools are available for identifying problems in a dataset, the activity of data cleaning still suffers from a challenge that has recieved increasing attention in the scientitic communities in the later years: Data cleaning is not very straight forward to document and therefore, reproducibility suffers. We present a new \proglang{R} package, \pkg{dataMaid} \citep{dataMaid}, whose most central purpose is to facilitate a data cleaning workflow where documentation is thoroughly integrated rather than a secondary add-on. This is accomplished by centering the data cleaning process around auto-generated data overview reports that summarize the contents of the current version of a dataset and flags potential problems that could be candidates for the next bullet on the cleaning agenda.

A number of \proglang{R} packages made for other pre-analysis steps are
already available, including \pkg{janitor} \citep{janitor}, \hl{assertive}, \pkg{dplyr}
\citep{dplyr}, \pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, \pkg{assertr} \citep{assertr}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \hl{Mention assertive here}. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work observation-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
variable-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process. The \pkg{assertr} package provide very similar - and very nice - tools to those of \pkg{validate}, but without any amibitions of conducting auto-cleaning.

@@

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
   var1 var2  var3       var4 var5       var6
1   red    1     a -0.6264538    1 Irrelevant
2   red    1     a  0.1836433    2 Irrelevant
3   red    1     a -0.8356286    3 Irrelevant
4   red    2     a  1.5952808    4 Irrelevant
5   red    2     a  0.3295078    5 Irrelevant
6   red    6     b -0.8204684    6 Irrelevant
7   red    6     b  0.4874291    7 Irrelevant
8   red    6     b  0.7383247    8 Irrelevant
9   red  999     c  0.5757814    9 Irrelevant
10  red   NA     c -0.3053884   10 Irrelevant
11 blue    4     c  1.5117812   11 Irrelevant
12 blue   82     .  0.3898432   12 Irrelevant
13 blue   NA       -0.6212406   13 Irrelevant
14 <NA>  NaN other -2.2146999   14 Irrelevant
15 <NA>    5 OTHER  1.1249309   15 Irrelevant
\end{Soutput}
\end{Schunk}

@@

By default, an \proglang{R} markdown file and a rendered pdf \hl{OBS: måske html/doc alt efter hvem der kalder funktionen?} overview
document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three parts, two of which 
are presented in Figure~\ref{fig:example1}. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, as shown in Figure~\ref{fig:example2}, for the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, output="html", render=FALSE, openResult=FALSE)
\end{Sinput}
\end{Schunk}

@@

$numeric
[1] "variableType" "countMissing" "uniqueValues" "centralValue" "quartiles"   
[6] "minMax"

$integer
[1] "variableType" "countMissing" "uniqueValues" "centralValue" "quartiles"   
[6] "minMax"

@@

$Date
[1] "variableType" "countMissing" "uniqueValues" "centralValue" "minMax"      
[6] "quartiles" 
\end{Soutput}
\end{Schunk} %$

@@

\begin{Schunk}
\begin{Sinput}
R> defaultCaracterSummaries()
\end{Sinput}
\begin{Soutput}
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
+    summaries = setSummaries(numeric = c("variableType","centralValue),
+    integer = c("variableType", "centralValue")))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
+    summaries = setSummaries(all = c("variableType", "centralValue")))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, 
+    checks = setChecks(factor = "identifyWhitespace"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, checks = setChecks(factor = NULL))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, preChecks = c("isKey", "isSupported"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, visuals = setVisuals(all = "basicVisual"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
>R complexData <- data.frame(complexVar = complex(100, real = 1:100, 
+    imaginary = 3), numericVar = 1:100))
>R makeDataReport(complexData, treatXasY(complex = "numeric"))
\end{Sinput}
\end{Schunk}

In this report, we will find that the two variables, \code{complexVar} and \code{numericVar} will be have identical presentations in the variable list, as treating a \code{complex} variable as a \code{numeric} means dropping the imaginary part of the complex numbers - which was the only thing setting the two variables apart in the first place.

@@

\subsection{Data cleaning by hand: An example}
Assume that we wish to look further into a certain variable from
\code{toyData}, namely \code{var2}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \code{check()} command

\begin{Schunk}
\begin{Sinput}
R> check(toyData$var2)
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
$identifyOutliers
Note that the following possible outlier values were detected: 82, 999.
\end{Soutput}
\end{Schunk}%$

@@

\begin{Schunk}
\begin{Sinput}
R> check(toyData$var2, checks = setChecks(numeric = "identifyMissing"))
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk} %$

@@

\begin{Schunk}
\begin{Sinput}
R> missVar2 <- identifyMissing(toyData$var2)
R> str(missVar2)
\end{Sinput}
\begin{Soutput}
List of 3
 $ problem      : logi TRUE
 $ message      : chr "The following suspected missing value codes
    enter as regular values: \\\"999\\\", \\\"NaN\\\"."
 $ problemValues: num [1:2] 999 NaN
 - attr(*, "class")= chr "checkResult"
\end{Soutput}
\end{Schunk}

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the entry
\code{problemValues}. If we decide that the values \code{999}
and \code{NaN} in \code{var2} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

\begin{Schunk}
\begin{Sinput}
R> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
No problems found.
\end{Soutput}
\end{Schunk} %$

Similarly, the \code{visualize()} and \code{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example2} for the visualization output.

\begin{Schunk}
\begin{Sinput}
R> visualize(toyData$var2)
R> summarize(toyData$var2)
\end{Sinput}
\begin{Soutput}
     Feature                   Result       
[1,] "Variable type"           "numeric"    
[2,] "Number of missing obs."  "4 (26.67 %)"
[3,] "Number of unique values" "6"          
[4,] "Median"                  "4"          
[5,] "1st and 3rd quartiles"   "1.5; 6"     
[6,] "Min. and max."           "1; 82"      
\end{Soutput}
\end{Schunk}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\caption{Output from running \code{visualize()} on the variable \code{var2} from the
\code{toyData} dataset.}
\label{fig:example2}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
R> summarize(toyData$var2, 
+    summaries = setSummaries(
+    numeric = defaultNumericSummaries(remove = c("variableType", 
+    "countMissing"))))
\end{Sinput}
\begin{Soutput}
     Feature         Result 
[1,] "Median"        "4"    
[2,] "Min. and max." "1; 82"
\end{Soutput}
\end{Schunk} %$

The syntax in this code chunk can be read as follows: "Summarize \code{var2} in \code{toyData}, and for \code{numeric} variables, set the summaries to be the default summary functions, except \code{variableType} and \code{countMissing}."

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData$lastName[presidentData$lastName == " Truman"] <- 
+    "Truman"
R> presidentData$stateOfBirth[presidentData$stateOfBirth == "New york"] <- 
+    "New York"
R> presidentData$assassinationAttempt <- 
+    factor(presidentData$assassinationAttempt)
R> presidentData$ageAtInauguration <- 
+    as.numeric(presidentData$ageAtInauguration)
\end{Sinput}
\end{Schunk}%$

@@

\begin{Schunk}
\begin{Sinput}
>R presidentData$ageAtInauguration <- 
+    as.numeric(as.character(presidentData$ageAtInauguration))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData$lastName[presidentData$firstName == "."]
\end{Sinput}
\begin{Soutput}
[1] "Trump"
\end{Soutput}
\begin{Sinput}
R> presidentData$firstName[presidentData$firstName == "."] <- "Donald"
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> birthdayOutlierVal <- 
+    identifyOutliers(presidentData$birthday)$problemValues
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> presidentData[presidentData$birthday == birthdayOutlierVal, ]
\end{Sinput}
\begin{Soutput}
      lastName firstName orderOfPresidency   birthday stateOfBirth
46 Arathornson   Aragorn                 0 1300-03-01       Gondor
   assassinationAttempt  sex ethnicity precidencyYears ageAtInauguration
46                    1 Male Caucasian              NA              88
   favoriteNumber
46           8+0i
\end{Soutput}
\end{Schunk}%$

@@

\begin{Schunk} 
\begin{Sinput}
R> presidentData <- 
+    presidentData[presidentData$birthday != birthdayOutlierVal, ]
\end{Sinput}
\end{Schunk} %$

@@

\begin{Schunk}
\begin{Sinput}
>R summarize(presidentData$presidencyYears)
\end{Sinput}
\begin{Soutput}
$variableType
Variable type: numeric
$countMissing
Number of missing obs.: 1 (2.22 %)
$uniqueValues
Number of unique values: 11
$centralValue
Median: 4
$quartiles
1st and 3rd quartiles: 4; 8
$minMax
Min. and max.: 0; Inf
\end{Soutput}
\end{Schunk} %$

@@

\begin{Schunk}
\begin{Sinput}
makeDataReport(presidentData, vol = "_cleaned", 
+    treatXasY = list(Name = "character", complex = "numeric"),
+    checks = setChecks(character = 
+    defaultCharacterChecks(remove = "identifyLoners")),
+    reportTitle = "Dirty president data - cleaned")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, onlyProblematic = TRUE, mode = "check")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> toyChecks <- check(toyData)
R> foo <- function(x) {
+    any(sapply(x, function(y) y[["problem"]]))
+  }
R> sapply(toyChecks, foo)
\end{Sinput}
\begin{Soutput}
 var1  var2  var3  var4  var5  var6 
 TRUE  TRUE  TRUE  TRUE  TRUE FALSE 
\end{Soutput}
\end{Schunk}

and we find that only a single variable in \code{toyData}, \code{var6} (for which all
observations have the value \code{"Irrelevant"}), is problem-free when using the default checks.

@@

Alternatively, one can create the usual output report, not render it and then manually edit the produced \proglang{R} markdown file as wished. This can be done with the following command:
\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, render = FALSE, openResult = FALSE)
\end{Sinput}
\end{Schunk}
After editing, the file can be rendered by calling the \code{render} function:
\begin{Schunk}
\begin{Sinput}
R> render("dataMaid_toyData.Rmd", quiet = FALSE)
\end{Sinput}
\end{Schunk}

>>>>

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should generally not
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need a subject
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the information needed for identifying issues.

But even when tools are available for identifying problems in a
dataset, the activity of data cleaning still suffers from a challenge
that has recieved increasing attention in the scientitic communities
in the later years: Data cleaning is not very straight forward to
document and therefore, reproducibility suffers. We present a new
\proglang{R} package, \pkg{dataMaid} \citep{dataMaid}, whose most
central purpose is to facilitate a supervised data quality screening
workflow where documentation is thoroughly integrated rather than an
add-on. This is accomplished by structuring the data screening around
auto-generated data overview reports that summarize and flags
potential problems in the dataset.

But no matter how clever software tools we make, data cleaning remains
to be a time consuming endeavor, which inherently requires human
interaction since every dataset is different and the variables in the
dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

The \pkg{dataMaid} approach to data screening, -quality assessment and
-documentation is governed by two fundamental paradigms. First of all,
there is no need for data cleaning to be an ad hoc procedure. Often,
we have a very clear idea of what flags are raisable in a given
dataset before we look at it, as we were the ones to produce it in the
first place. This means that data cleaning can easily be a
well-documented, well-specified procedure. In order to aid this
paradigm, \pkg{dataMaid} provides easy-to-use, automated tools for
data quality assessment in \proglang{R} \citep{R} on which data
cleaning decisions can be made. This quality assessment is presented
in an auto-generated overview document, readable by data analysts and
field experts alike, thereby also contributing to an inter-field
dialogue about the data at hand. Oftentimes, e.g., distinguishing
between faulty codings of a numeric value and unusual, but correct,
values requires problem-specific expertise that might not be held by
the data analyst. Hopefully, having easy access to data summaries
through \pkg{dataMaid} will help this necessary knowledge sharing.

A number of \proglang{R} packages made for other pre-analysis steps
are already available, including \pkg{janitor} \citep{janitor},
\pkg{assertive} \citep{assertive}, \pkg{dplyr} \citep{dplyr},
\pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table},
\pkg{DataCombine} \citep{DataCombine}, \pkg{validate}
\citep{validate}, \pkg{assertr}, and \citep{assertr}. These packages
focus on different stages of the pre-analysis work.  \pkg{janitor}
provides tools for data import with a particular emphasis on the
challenges of getting neat data frames from Microsoft Excel data
files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table} and
\pkg{DataCombine} go a few steps further by providing a wide array of
extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) and \pkg{assertive} offers tools for identifying errors
in a dataset by checking the state of the variable given a set of
pre-specified rules, and their focus is on internal validity rather
than general data screening. In practice, this means that quite elegant
tools for, e.g., linear restraints among the variables in a dataset
can be applied, but looking for potentially miscoded missing values is
not really feasible. The main difference between these two challenges
is the direction in which the data is inspected: While linear
constraints work observation-wise with no ambiguity, determining
whether or not something is a miscoded missing value often requires
knowledge about the full variable (e.g. range or data type), and thus
it should be performed variable-wise. \pkg{validate} does not
currently allow for user-defined extensions of the latter type,
thereby limiting its data cleaning potential.  Automatic data
correction functions are also provided by \pkg{validate} which we
consider to be quite a dangerous cocktail: all power is given to the
the computer with no human supervision, and investigators are less
likely to make an active, case-specific choice regarding the handling
of the potential errors.  Finally, no tools have been made to easily
document exactly which checks and preliminary results were used in the
data cleaning process. The \pkg{assertr} package provide very similar
--- and very nice --- tools to those of \pkg{validate}, but without
any amibitions of conducting auto-cleaning.

@@

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
# A tibble: 15 x 6
     pill events region     change     id spotifysong
   <fctr>  <dbl> <fctr>      <dbl> <fctr>      <fctr>
 1    red      1      a -0.6264538      1  Irrelevant
 2    red      1      a  0.1836433      2  Irrelevant
 3    red      1      a -0.8356286      3  Irrelevant
 4    red      2      a  1.5952808      4  Irrelevant
 5    red      2      a  0.3295078      5  Irrelevant
 6    red      6      b -0.8204684      6  Irrelevant
 7    red      6      b  0.4874291      7  Irrelevant
 8    red      6      b  0.7383247      8  Irrelevant
 9    red    999      c  0.5757814      9  Irrelevant
10    red     NA      c -0.3053884     10  Irrelevant
11   blue      4      c  1.5117812     11  Irrelevant
12   blue     82      .  0.3898432     12  Irrelevant
13   blue     NA        -0.6212406     13  Irrelevant
14   <NA>    NaN  other -2.2146999     14  Irrelevant
15   <NA>      5  OTHER  1.1249309     15  Irrelevant
\end{Soutput}
\end{Schunk}

@@

By default, an \proglang{R} markdown file and a rendered pdf \hl{OBS: måske html/doc alt efter hvem der kalder funktionen?} overview
document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three parts, two of which 
are presented in Figure~\ref{fig:example1}. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, as shown in Figure~\ref{fig:example2}, for the \code{numeric}-type variable
\code{events} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, output="html", render=FALSE, openResult=FALSE, replace=TRUE)
\end{Sinput}
\end{Schunk}

@@

$numeric
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
[5] "quartiles"    "minMax"

$integer
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
[5] "quartiles"    "minMax"

@@

$Date
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
[5] "minMax"       "quartiles"   
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> defaultCharacterSummaries()
\end{Sinput}
\begin{Soutput}
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, replace=TRUE,
+    summaries = setSummaries(numeric = c("variableType","centralValue"),
+    integer = c("variableType", "centralValue")))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, replace=TRUE,                               
+    summaries = setSummaries(all = c("variableType", "centralValue")))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, replace=TRUE,
+    checks = setChecks(factor = "identifyWhitespace"))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, checks = setChecks(factor = NULL), replace=TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, preChecks = c("isKey", "isSupported"), replace=TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, visuals = setVisuals(all = "basicVisual"), replace=TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> complexData <- data.frame(complexVar = complex(100, real = 1:100, imaginary = 3), numericVar = 1:100)
R> makeDataReport(complexData, treatXasY=list(complex = "numeric"), replace=TRUE)
\end{Sinput}
\end{Schunk}

In this report, we will find that the two variables, \code{complexVar} and \code{numericVar} will be have identical presentations in the variable list, as treating a \code{complex} variable as a \code{numeric} means dropping the imaginary part of the complex numbers which was the only thing setting the two variables apart in the first place.

@@

\subsection{Data cleaning by hand: An example}
Assume that we wish to look further into a certain variable from
\code{toyData}, namely \code{events}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \code{check()} command

\begin{Schunk}
\begin{Sinput}
R> check(toyData$events)
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
$identifyOutliers
Note that the following possible outlier values were detected: 82, 999.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> check(toyData$events, checks = setChecks(numeric = "identifyMissing"))
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> identifyMissing(toyData$events)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> missEvents <- identifyMissing(toyData$events)
R> str(missEvents)
\end{Sinput}
\begin{Soutput}
List of 3
 $ problem      : logi TRUE
 $ message      : chr "The following suspected missing value codes enter as regular values: \\\"999\\\", \\\"NaN\\\"."
 $ problemValues: num [1:2] 999 NaN
 - attr(*, "class")= chr "checkResult"
\end{Soutput}
\end{Schunk}

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the entry
\code{problemValues}. If we decide that the values \code{999}
and \code{NaN} in \code{events} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

\begin{Schunk}
\begin{Sinput}
R> toyData$events[toyData$events %in% missEvents$problemValues] <- NA
R> identifyMissing(toyData$events)
\end{Sinput}
\begin{Soutput}
No problems found.
\end{Soutput}
\end{Schunk}

Similarly, the \code{visualize()} and \code{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example3} for the visualization output.

\begin{Schunk}
\begin{Sinput}
R> visualize(toyData$events)
R> summarize(toyData$events)
\end{Sinput}
\begin{Soutput}
$variableType
Variable type: numeric
$countMissing
Number of missing obs.: 4 (26.67 %)
$uniqueValues
Number of unique values: 6
$centralValue
Median: 4
$quartiles
1st and 3rd quartiles: 1.5; 6
$minMax
Min. and max.: 1; 82
\end{Soutput}
\end{Schunk}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-events.pdf}
\end{center}
\caption{Output from running \code{visualize()} on the variable \code{events} from the
\code{toyData} dataset.}
\label{fig:example3}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
R> summarize(toyData$events, 
+    summaries = setSummaries(
+    numeric = defaultNumericSummaries(remove = c("variableType", 
+    "countMissing"))))
\end{Sinput}
\begin{Soutput}
$uniqueValues
Number of unique values: 6
$centralValue
Median: 4
$quartiles
1st and 3rd quartiles: 1.5; 6
$minMax
Min. and max.: 1; 82
\end{Soutput}
\end{Schunk}

The syntax in this code chunk can be read as follows: "Summarize \code{events} in \code{toyData}, and for \code{numeric} variables, set the summaries to be the default summary functions, except \code{variableType} and \code{countMissing}."

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData$lastName[presidentData$lastName == " Truman"] <- "Truman"
R> presidentData$stateOfBirth[presidentData$stateOfBirth == "New york"] <- "New York"
R> presidentData$assassinationAttempt <- factor(presidentData$assassinationAttempt)
R> presidentData$ageAtInauguration <- as.numeric(presidentData$ageAtInauguration)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData$ageAtInauguration <- as.numeric(as.character(presidentData$ageAtInauguration))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData$lastName[presidentData$firstName == "."]
\end{Sinput}
\begin{Soutput}
[1] "Trump"
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
R> presidentData$firstName[presidentData$firstName == "."] <- "Donald"
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> birthdayOutlierVal <- 
+  +    identifyOutliers(presidentData$birthday)$problemValues
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData[presidentData$birthday == birthdayOutlierVal, ]
\end{Sinput}
\begin{Soutput}
      lastName firstName orderOfPresidency   birthday stateOfBirth
46 Arathornson   Aragorn                 0 1300-03-01       Gondor
   assassinationAttempt  sex ethnicity presidencyYears
46                    1 Male Caucasian              NA
   ageAtInauguration favoriteNumber
46                87           8+0i
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> presidentData <- 
+    presidentData[presidentData$birthday != birthdayOutlierVal, ]
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> summarize(presidentData$presidencyYears)
\end{Sinput}
\begin{Soutput}
$variableType
Variable type: numeric
$countMissing
Number of missing obs.: 1 (2.22 %)
$uniqueValues
Number of unique values: 10
$centralValue
Median: 4
$quartiles
1st and 3rd quartiles: 3.75; 8
$minMax
Min. and max.: 0; Inf
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(presidentData, vol = "_cleaned", 
+    treatXasY = list(Name = "character", complex = "numeric"),
+    checks = setChecks(character = 
+    defaultCharacterChecks(remove = "identifyLoners")),
+    reportTitle = "Dirty president data - cleaned",
+    replace=TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, onlyProblematic = TRUE, mode = "check", replace=TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> toyChecks <- check(toyData)
R> foo <- function(x) {
+    any(sapply(x, function(y) y[["problem"]]))
+  }
R> sapply(toyChecks, foo)
\end{Sinput}
\begin{Soutput}
       pill      events      region      change          id 
       TRUE        TRUE        TRUE        TRUE        TRUE 
spotifysong 
      FALSE 
\end{Soutput}
\end{Schunk}

and we find that only a single variable in \code{toyData}, \code{spotifysong} (for which all
observations have the value \code{"Irrelevant"}), is problem-free when using the default checks.

@@

Alternatively, one can create the usual output report, not render it and then manually edit the produced \proglang{R} markdown file as wished. This can be done with the following command:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, render = FALSE, openResult = FALSE)
\end{Sinput}
\end{Schunk}

After editing, the file can be rendered by calling the \code{render} function:

\begin{Schunk}
\begin{Sinput}
R> render("dataMaid_toyData.Rmd", quiet = FALSE)
\end{Sinput}
\end{Schunk}
